Qualitative Probabilistic Programming

### 2015-06-05 paper/paper.tex
paperrr

\documentclass{article}
\usepackage{cite}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{array}
% \usepackage{minted1}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{multicol} \usepackage{fancyheadings} \usepackage{pdfpages}
\usepackage{nips15submit_e,times}
\setlength{\emergencystretch}{10em}

@@

  In an ordinary probabilistic programming language (such as Church),
  it is possible to treat parameters as random variables.  This
  would allow ordinary inference algorithms to infer parameters.  However,
  there are advantages of having unknown functions as a feature
  in the language.
  First, it is easier to
  write programs without knowing the details of different parameterized distributions.
  Second, the system can use specialized algorithms to infer parameters faster.

@@

  \begin{lstlisting}
var Cluster = Categorical(2);
var Point = Vector(2, Double);
var getPoint = randFunction(Cluster, Point);

var model = function() {
  var cluster = randomValue(Cluster);
  observe(getPoint, cluster);
};
  \end{lstlisting}

@@

  To demonstrate, let us run this example on a dataset consisting of 150 points (TODO cite).  When we run the program on this data, we infer the parameters to the random function \texttt{getPoint}.
  In this case, \texttt{getPoint} is a linear function with Gaussian noise, so it will naturally
  split the data into 2 clusters with equal variance:
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_orig.png}
  \end{center}

  The first cluster is at (6.3, 5.0) and the second is at (5.0, 1.6).  They both
  have a standard deviation of 0.54 in the x direction and 0.69 in the y direction.
  We could use these parameters to fill in the generative model, giving us a webppl program:
  \begin{lstlisting}
var model = function() {
  var cluster = randomInteger(nclusters);
  return [gaussian(cluster == 0 ? 6.3 : 5.0, 0.54),
          gaussian(cluster == 0 ? 5.0 : 1.6, 0.69)];
};
  \end{lstlisting}
  This model is estimated to assign probability density $e^{-393}$ to the data, yielding a perplexity of $e^{-393/150} = 0.0728$.

@@

  In the previous example, we that \texttt{randFunction(Categorical(2), Vector(2, Double))} represented 2 axis-aligned 2-dimensional Gaussian clusters with equal variances.  In general, \texttt{randFunction} returns a randomized function that is a member of some generalized linear model
    determined by the desired argument types and return type.
    The distribution
    of the function's return value $y$
    is some exponential family whose natural
    parameters are determined from the arguments $x$:

@@

    Additionally, we provide a $\texttt{randomValue}$ function for convenience, which
    returns a sample from the ``default'' distribution for a given type.  The following shows the types
    in Quipp and their corresponding sufficient statistics, features, and English descriptions.

    \begin{center}
    {\scriptsize
      \begin{tabular}{| c | c | c | c | C{3cm} |}
      \hline
      T & $\phi_T(X)$ & $\psi_T(X)$ & \texttt{randomValue(T)} & Random function class \\

@@

      \hline
    \texttt{Tuple(T1, T2)} & $\begin{bmatrix} \phi_{T1}(X_1) \\ \phi_{T2}(X_2) \end{bmatrix}$
                           & $\begin{bmatrix} \psi_{T1}(X_1) \\ \psi_{T2}(X_2) \end{bmatrix}$ &
      \begin{tabular}{l}
        \texttt{[randomValue(T1),}
       \\
     \texttt{ randomValue(T2)]}
      \end{tabular}
                           &
      Independent regression
      \\
      \hline
    \end{tabular}
  }
\end{center}

@@

  \section{Example: improving the clustering model}
  Using this knowledge about the form of the random functions, we can improve the clustering model from before.
  As observed in the previous graph, the two clusters are forced to have the
  same variance. They do not fit the data well, since the data has a different
  shape in each location.  To fix this problem, we can substitute the following
  model, which uses a separate random function (and therefore a separate axis-aligned Gaussian distribution) for each cluster:
  \begin{lstlisting}
var Cluster = Categorical(2);
var Point = Vector(2, Double);
var getPointFunctions = [randFunction(Point), randFunction(Point)];

var model = function() {
  var cluster = randomValue(Cluster);
  observe(getPointFunctions[cluster]);
};
\end{lstlisting}
  Using this model, we get the following clusters:
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_indep.png}
  \end{center}
  This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, yielding a perplexity of 0.1058.  This is a significantly improved fit.

@@

  Here, there are 2 random functions for each cluster, one to get the $x$ coordinate and one to get the $y$ coordinate (whose distribution may depend linearly on the $x$ coordinate).  This allows $x$ and $y$ to be correlated, improving fit:

  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_dep.png}
  \end{center}

  This model is estimated to assign probability $e^{-269}$ to the data, yielding a perplexity of 0.166.  This is a large improvement from the previous model.

@@

    To evaluate performance, for each model, we:
    \begin{itemize}
      \item
        Randomly generate parameters $\theta$
      \item
        Generate datasets $x_{train}, x_{test}$ using $\theta$
      \item
        Estimate $\log P(x_{test} | \theta)$
      \item
        Use the EM algorithm to infer approximate parameters $\hat{\theta}_i$ from $x_{train}$, for iteration $i = 1, 2, ...$
      \item
        Estimate $\log P(x_{test} | \hat{\theta}_i)$
      \item
        Use the previous estimates to estimate the regret $\log P(x_{test} | \hat{\theta}_i) - \log P(x_{test} | \theta)$
    \end{itemize}
    Estimating $\log P(x_{test} | \theta)$ is nontrivial, given that the model contains latent variables.
    We use the Sequential Monte Carlo algorithm for this, as described in \cite{dippl}.

@@

\begin{center}
  \includegraphics[scale=0.5]{../plots/accuracy_nd_clustering.png}
\end{center}

@@

var model = function() {
  var whichClass = randomValue(Class);
  observe(classFeatures, whichClass);
};
\end{lstlisting}
}

The naive Bayes model is similar to the clustering model.  We have two classes and a feature
distribution for each.  Since each feature is boolean, we will learn
a different categorical distribution for each class.

(figure should show average classification accuracy)

@@

\begin{center}
  \includegraphics[scale=0.5]{../plots/accuracy_factor_analysis.png}
\end{center}

@@

\begin{center}
  \includegraphics[scale=0.5]{../plots/accuracy_hmm.png}
\end{center}

@@

\begin{center}
  \includegraphics[scale=0.5]{../plots/accuracy_lda.png}
\end{center}
\subsection{Neural network}

@@

var model = function() {
  var inputLayer = randomValue(Input);
  var hiddenLayer = inputToHidden(inputs[sampIndex]);
  observe(hiddenToOutput, hiddenLayer);
};
\end{lstlisting}
}

@@

  In the future, it will be useful to support additional types.  For example, we could add types for standard exponential families such as Poisson and Gamma.  Supporting disjoint union types is also straightforward.  Also, it will be useful to create a more usable interface to infer parameters and perform additional data processing given these parameters.

>>>>

\documentclass{article}
\usepackage{cite}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{placeins}
\usepackage{array}
% \usepackage{minted1}
\usepackage{hyperref}
\usepackage{amsfonts}
\usepackage{multicol} \usepackage{fancyheadings} \usepackage{pdfpages}
\usepackage{nips15submit_e,times}
\setlength{\emergencystretch}{10em}

@@

  % In an ordinary probabilistic programming language (such as Church),
  % it is possible to treat parameters as random variables.  This
  % would allow ordinary inference algorithms to infer parameters.  However,
  % there are advantages of having unknown functions as a feature
  % in the language.
  % First, it is easier to
  % write programs without knowing the details of different parameterized distributions.
  % Second, the system can use specialized algorithms to infer parameters faster.

@@

  {
  \begin{lstlisting}
var Cluster = Categorical(2);
var Point = Vector(2, Double);
var getPoint = randFunction(Cluster, Point);

@@

  To demonstrate, let us run this example on the Iris dataset, which consists of 150 points \cite{Lichman:2013}.  When we run the program on this data, we infer the parameters to the random function \texttt{getPoint}.
  In this case, \texttt{getPoint} is a linear function with Gaussian noise, so it will naturally
  split the data into 2 clusters with equal variance, as seen in Figure 1.

  \begin{figure}[h]
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_orig.png}
  \end{center}
  \caption{Clustering the Iris dataset}
\end{figure}

  The first cluster is at (6.3, 5.0) and the second is at (5.0, 1.6).  They both
  have a standard deviation of 0.54 in the x direction and 0.69 in the y direction.
  We could use these parameters to fill in the generative model, giving us a webppl program:
  {
  \begin{lstlisting}
var model = function() {
  var cluster = randomInteger(nclusters);
  return [gaussian(cluster == 0 ? 6.3 : 5.0, 0.54),
          gaussian(cluster == 0 ? 5.0 : 1.6, 0.69)];
};
  \end{lstlisting}
}
  This model is estimated to assign probability density $e^{-393}$ to the data, yielding a perplexity of $e^{-393/150} = 0.0728$.

@@

  In the previous example, \texttt{randFunction(Categorical(2), Vector(2, Double))} represented 2 unknown axis-aligned 2-dimensional Gaussian clusters with equal variances.  In general, \texttt{randFunction} returns a randomized function that is a member of some generalized linear model
    determined by the desired argument types and return type.
    The distribution
    of the function's return value $y$
    is some exponential family whose natural
    parameters are determined from the arguments $x$:

@@

    Additionally, we provide a $\texttt{randomValue}$ function for convenience, which
    returns a sample from the ``default'' distribution for a given type.  Table 1 shows the types
    in Quipp and their corresponding sufficient statistics, features, and English descriptions.

\begin{table}[t]
  \caption{Types in Quipp}
  \label{type-table}
    \begin{center}
    {\scriptsize
      \begin{tabular}{| c | c | c | c | C{3cm} |}
      \hline
      T & $\phi_T(X)$ & $\psi_T(X)$ & \texttt{randomValue(T)} & Random function class \\

@@

      \hline
    \texttt{Tuple(T1, T2)} & $\begin{bmatrix} \phi_{T1}(X_1) \\ \phi_{T2}(X_2) \end{bmatrix}$
                           & $\begin{bmatrix} \psi_{T1}(X_1) \\ \psi_{T2}(X_2) \end{bmatrix}$ &
      \begin{tabular}{l}
        \texttt{[randomValue(T1),}
       \\
     \texttt{ randomValue(T2)]}
      \end{tabular}
                           &
      Independent regression
      \\
      \hline
    \end{tabular}
  }
\end{center}
\end{table}

@@

  \section{Example: improving the clustering model}
  Using this knowledge about the form of the random functions, we can improve the clustering model from before.
  As observed in Figure 1, the two clusters are forced to have the
  same variance. They do not fit the data well, since the data has a different
  shape in each location.  To fix this problem, we can substitute the following
  model, which uses a separate random function (and therefore a separate axis-aligned Gaussian distribution) for each cluster:
  \begin{lstlisting}
var Cluster = Categorical(2);
var Point = Vector(2, Double);
var getPointFunctions = [randFunction(Point), randFunction(Point)];

var model = function() {
  var cluster = randomValue(Cluster);
  observe(getPointFunctions[cluster]);
};
\end{lstlisting}
  \begin{figure}[h]
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_indep.png}
  \end{center}
  \caption{Clustering the Iris dataset with independent variances}
\end{figure}
  Using this model, we get the clusters shown in Figure 2.
  This model (with the parameters filled in) is estimated to assign probability density $e^{-337}$ to the data, yielding a perplexity of 0.1058.  This is a significantly improved fit.

@@

  Here, there are 2 random functions for each cluster, one to get the $x$ coordinate and one to get the $y$ coordinate (whose distribution may depend linearly on the $x$ coordinate).  This allows $x$ and $y$ to be correlated, improving fit, as seen in Figure 3.
  This model is estimated to assign probability $e^{-269}$ to the data, yielding a perplexity of 0.166.  This is a large improvement from the previous model.

  \begin{figure}[h]
  \begin{center}
    \includegraphics[scale=0.5]{../plots/irisclusters_dep.png}
  \end{center}
  \caption{Clustering the Iris dataset with multivariate Gaussian distributions}
\end{figure}

@@

    To evaluate performance, for each model, we:
    \begin{itemize}
      \item
        Randomly generate parameters $\theta$
      \item
        Generate datasets $x_{train}, x_{test}$ using $\theta$
      \item
        Estimate $\log P(x_{test} | \theta)$
      \item
        Use the EM algorithm to infer approximate parameters $\hat{\theta}_i$ from $x_{train}$, for iteration $i = 1, 2, ...$
      \item
        Estimate $\log P(x_{test} | \hat{\theta}_i)$
      \item
        Use the previous estimates to estimate the regret $\log P(x_{test} | \hat{\theta}_i) - \log P(x_{test} | \theta)$
    \end{itemize}
    Estimating $\log P(x_{test} | \theta)$ is nontrivial, given that the model contains latent variables.
    We use the Sequential Monte Carlo algorithm for this, as described in \cite{dippl}.  For the following examples,
    for each iteration except the first, we plot quantiles of average regret per sample over multiple runs.  We exclude
    the first iteration because it represents the regret of randomly generated parameters, which is much higher than the regret in future iterations.  Since
    the expectation maximization algorithm does not always converge to a global optimum, we do not expect regret
    to go to 0 as we increase the number of iterations.

@@

  \begin{figure}[h]
\begin{center}
  \includegraphics[scale=0.5]{../plots/accuracy_nd_clustering.png}
\end{center}
\caption{Regret-per-sample quantiles for clustering}
\end{figure}

@@

var model = function() {
  var whichClass = observe(randomValue, Class);
  observe(classFeatures[whichClass]);
};
\end{lstlisting}
}

The naive Bayes model is similar to the clustering model.  We have two classes and a feature
distribution for each.  Since each feature is boolean, we will learn
a different categorical distribution for each class.  The main difference from clustering is that,
since this is a supervised learning method, we observe the class in addition to the features.  Since there
are no latent variables, the system finds the correct parameters in a single iteration, so the regret plot is uninteresting.

@@

  \begin{figure}[h]
\begin{center}
  \includegraphics[scale=0.5]{../plots/accuracy_factor_analysis.png}
\end{center}
\caption{Regret-per-sample quantiles for factor analysis}
\end{figure}

@@

  \begin{figure}[h]
\begin{center}
  \includegraphics[scale=0.5]{../plots/accuracy_hmm.png}
\end{center}
\caption{Regret-per-sample quantiles for hidden Markov model}
\end{figure}

@@

  \begin{figure}[h]
\begin{center}
  \includegraphics[scale=0.5]{../plots/accuracy_lda.png}
\end{center}
\caption{Regret-per-sample quantiles for latent Dirichlet allocation}
\end{figure}
\subsection{Neural network}

@@

var model = function() {
  var inputLayer = observe(randomValue, Input);
  var hiddenLayer = inputToHidden(inputs[sampIndex]);
  observe(hiddenToOutput, hiddenLayer);
};
\end{lstlisting}
}

@@

  In the future, it will be useful to support additional types.  For example, we could add types for standard exponential families such as Poisson and Gamma.  Supporting disjoint union types is also straightforward.  Also, it will be useful to create a more user-friendly interface for inferring parameters and performing additional data processing, such as inferring latent values given fixed parameters or visualizing the distributions implied by the parameters.
